\normalfont \LARGE Improved Quasi-Static Methods for Time-Dependent Neutron Diffusion and Implementation in Rattlesnake

### 2016-09-28 doc/Thesis_proposal/thesis_proposal.tex
jcr comments to the ms proposal

\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand{\vn}{\vec{n}}
\newcommand{\vel}{\vec{\mathrm{v}}}
\newcommand{\adj}{\Phi^\dagger_0}

@@

Transient modeling of nuclear reactors has been a chronically foreboding task due to its computationally expensive nature.  However, recent developments in nuclear testing, including the revitalization of the Transient Reactor Testing (TREAT) Facility at Idaho National laboratory (INL), have brought significant attention to the development of transient reactor modeling.  Transient computational methods currently in laboratory frameworks have proven to be overbearingly slow for complex, real-world scenarios.  Therefore, more efficient and faster methods for computational frameworks are highly desired.  The Department of Energy (DOE) has especially sought to develop in the MOOSE framework.  \\

@@

The difficulty in transient reactor modeling is due to the neutron transport equation and the complicated nature of neutron motion that the equation represents.  The equation has seven independent variables: space ($\vec{r}$), energy ($E$), direction ($\vec{\Omega}$), and time ($t$) \cite{duderstadt1976nuclear}.  Nuclear reactor simulations often utilize the neutron diffusion approximation, which carefully eliminates the dependence on direction.  For the purposes of this research, neutron diffusion will be discussed exclusively.  Additionally, simulations can often neglect temporal dependence due to the steady-state nature of reactors.  However, simulations for facilities such as TREAT, must retain time dependence in order to obtain relevant information. \\

For computational purposes, the neutron diffusion equation is evaluated with discretized space, energy, and time.  There are many discretization schemes for each of these variables, but FEM for space and multigroup method for energy were used for the development of this research.  Since the purpose of this research involves temporal dependence, discretization in time was kept general.  The time dependent neutron diffusion equation with delayed neutron precursors can be seen in Equations~\ref{eq:flux}~and~\ref{eq:precursor}.

@@

Where,

@@

Most reactor computation frameworks, including Rattlesnake, discretize the time variable directly with these equations using a multitude of schemes (Implicit Euler, Crank-Nicholson, Runge-Kutta, etc.).  In this paper, this method of discretizing Equations~\ref{eq:flux}~and~\ref{eq:precursor} is considered brute-force method.  This research intends to improve upon this method by instead implementing the improved quasi-static method (IQS). \\

@@

This section describes the theory of IQS and the current status of its development.  The theory and status are broken up into three section: \sct{sect:theory} analytically derives the method, \sct{sect:iter} and \sct{sect:pc} present two solution techniques and describes their application and development.

@@

We note that the time-dependent shape equation is similar to the time-dependent flux, with the following modifications:
\begin{enumerate}
\item The shape equation contains an additional term equivalent to a removal cross section,  $ \frac{1}{v^g}\frac{1}{p}\frac{dp}{dt}$.
\item The delayed neutron source term is divided by $p$.
\item The system of equations is now nonlinear due to the factorization.
\item An equation is needed to obtain the amplitude $p$.
\end{enumerate}

To derive the amplitude equation, the shape/precursors equations are weighted by a space-dependent function and integrated over the phase-space. The weight function is typically the adjoint flux $\phi^{*g}$, which can be proven to minimize truncation error \cite{duderstadt1976nuclear}.  The final expressions are given below:
\begin{subequations}
\be
\frac{dp}{dt}=\left[\frac{\rho-\bar{\beta}}{\Lambda}\right]p+\sum_{i=1}^I\bar{\lambda}_i\xi_i
\label{eq:p}
\ee
\be
\frac{d\xi_i}{dt}=\frac{\bar{\beta}_i}{\Lambda}-\bar{\lambda}_i\xi_i \quad 1 \le i \le I 
\label{eq:c}
\ee
\end{subequations}
This equation is also known as the point reactor kinetics equation (PRKE); where the reactivity, effective delayed-neutron fraction, and delayed-neutron precursor decay constant are defined as follows:
\begin{subequations}
\be
\frac{\rho-\bar{\beta}}{\Lambda} = 
\frac{ \sum_{g=1}^G\left(\phi^{*g},\frac{\chi_p^g}{\keff}(1-\beta)\sum_{g'=1}^G \nu^{g'} \Sigma_f^{g'}\varphi^{g'} + \sum_{g'\neq g}^G\Sigma_s^{g'\to g} \varphi^{g'} -\left( -\div D^g \grad  + \Sigma_r^g \right)\varphi^g\right)}
{\sum_{g=1}^G\left(\phi^{*g},\frac{1}{v^g}\varphi^g\right)}
\label{eq:rmb}
\ee
\be
\frac{\bar{\beta}}{\Lambda} = \sum_{i=1}^I\frac{\bar{\beta}_i}{\Lambda} = 
\sum_{i=1}^I\frac{1}{\keff}\frac{\sum_{g=1}^G(\phi^{*g}, \chi_{d,i}^g\beta_i\sum_{g'=1}^G\nu^{g'} \Sigma_f^{g' }\varphi^{g'})}
{\sum_{g=1}^G\left(\phi^{*g},\frac{1}{v^g}\varphi^g\right)}
\label{eq:b}
\ee
\be
\bar{\lambda}_i = \frac{\sum_{g=1}^G(\phi^{*g},\chi_{d,i}^g\lambda_i C_i)}{\sum_{g=1}^G(\phi^{*g},\chi_{d,i}^gC_i)}
\label{eq:l}
\ee 
\end{subequations}
The following inner product definition has been used: $\left(\phi^{*g},f\right):=\int_D\phi^{*g}(\vec{r})f(\vec{r})d^3r$.\\

@@

Solving for the shape in \eqt{eq:shape} can become expensive, especially in two or three dimensions, and even more so when using the transport equations in lieu of the diffusion equations.  Using IQS, one expects the time dependence of the shape to be weaker than that of the flux itself,  thus allowing for larger time step sizes in updating the shape. The PRKE equations form a small ODE system and can be solved using a much smaller time step size. In transients where the shape varies much less than the flux, IQS can thus be very computationally effective. The two time scale solution process, a micro scale for the PRKE and a macro scale for the shape, are illustrated in \fig{fig:iqsviz}.

@@

The major difference between the methods of these authors is the convergence criteria used.  Sissaoui and Koclas \cite{Sissaoui_1995, Koclas_1996} use fixed point iteration where the criteria is the simply the normalized difference between the last two computed shapes.  Monier in \cite{Monier_diss} also does fixed point iteration with the same criteria, except the solution is scaled by $\frac{\sum_{g=1}^G\left(\phi^{*g},\frac{1}{v^g}\varphi^g(t_{n})\right)}{\sum_{g=1}^G\left(\phi^{*g},\frac{1}{v^g}\varphi^g(t_{n+1})\right)}$ after each iteration.  Devooght in \cite{Devooght_1984} does a Newton-SOR iteration where the residual of the shape function evaluation is the convergence criteria and next iteration's solution is computed using Newton-Raphson method.\\

@@

The Predictor-Corrector (P-C) version of IQS factorizes the flux and derives the PRKE the same way as the standard version, but the solution of the coupled system of equations is different.  In the IQS P-C version, the flux equations (not the shape equations) are solved (represented by Equations~\ref{eq:flux}~and~\ref{eq:precursor}) in order to obtain a predicted flux solution. This predicted flux is then converted to shape by rescaling as follows:
\be
\varphi^g_{n+1} = \underbrace{\phi^g_{n+1}}_{\text{predicted}} \frac{K_0}{K_{n+1}}
\label{eq:rescale}
\ee
where the scaling factors are given by
\be
K_{n+1} =\sum_{g=1}^G\left(\phi^{*g},\frac{1}{v^g}\phi^g_{n+1}\right)
\ee
\be
K_{0} =\sum_{g=1}^G\left(\phi^{*g},\frac{1}{v^g}\varphi^g_{n+1}\right)=\sum_{g=1}^G\left(\phi^{*g},\frac{1}{v^g}\phi^g_{0} \,.\right)
\ee

@@

@@

A one-dimensional, single group FEM prototype will be created for IQS in order to test validation for simple problems. However, testing complex, real-world problems is most easily performed by implementing IQS in a FEM framework.  The Rattlesnake radiation transport framework is a excellent option for performing these problems.  Rattlesnake already contains a neutron transport method for CFEM diffusion equivalent to the brute-force method.  Additionally, Rattlesnake is well documented and well organized, so implementing IQS is relatively straightforward. With this implementation, benchmark testing such as 16-A1 \cite{ANL_BPB}, TWIGL, LRA, C5G7-2D, etc. can be performed, as well as a full core simulation of TREAT.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Other Possible Outlooks}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand{\vn}{\vec{n}}
\newcommand{\vel}{\vec{\mathrm{v}}}
\newcommand{\adj}{\Phi^\dagger_0}
\newcommand{\tcr}[1]{\textcolor{red}{#1}}

@@

Transient modeling of nuclear reactors has been a chronically foreboding task due to its computationally expensive nature.  However, recent developments in nuclear testing, including the revitalization of the Transient Reactor Testing (TREAT) Facility at Idaho National laboratory (INL), have brought significant attention to the development of transient reactor modeling.  Transient computational methods currently in \tcr{laboratory frameworks--people may not understand what you mean here} have proven to be overbearingly \tcr{slow-CPU-expensive} for complex, real-world scenarios.  Therefore, more efficient and faster methods for computational \tcr{frameworks} are highly desired.  The Department of Energy (DOE) has especially sought to develop in the MOOSE framework.  \\

@@

The difficulty in transient reactor modeling is due to the neutron transport equation and \tcr{the complicated nature of neutron motion that the equation represents-- ???? it is the dimensionality of the phase-space that is the issue, and the fact that the time discretization has to be implicit, which  leads to kinear system solves}.  The equation has seven independent variables: space ($\vec{r}$), energy ($E$), direction ($\vec{\Omega}$), and time ($t$) \cite{duderstadt1976nuclear}.  Nuclear reactor simulations often utilize the neutron diffusion approximation, which carefully eliminates the dependence on direction.  For the purposes of this research, neutron diffusion will be discussed exclusively.  \tcr{Additionally, simulations can often neglect temporal dependence due to the steady-state nature of reactors--???}.  However, simulations for facilities such as TREAT, must retain time dependence in order to obtain relevant information. \\

For computational purposes, the neutron diffusion equation is discretized in space, energy, and time.  There are many discretization schemes for each of these variables, but FEM for space and multigroup method for energy were used for the development of this research \tcr{I do not know of ``many'' discretizations in the energy variable ...}.  Since the purpose of this research involves temporal dependence, discretization in time was kept general.  The time dependent neutron diffusion equation with delayed neutron precursors can be seen in Equations~\ref{eq:flux}~and~\ref{eq:precursor}.

@@

where,

@@

Most reactor computation frameworks, including Rattlesnake, discretize the time variable directly with these equations using a multitude of schemes (Implicit Euler, Crank-Nicholson, Runge-Kutta, etc.).  In this paper, this method of discretizing Equations~\ref{eq:flux}~and~\ref{eq:precursor} is considered \tcr{brute-force--I wonder if we should come up with a better terminology...} method.  This research intends to improve upon this method by instead implementing the improved quasi-static method (IQS). \\

@@

This section describes the theory of IQS and the current status of its development.  The theory and status are broken up into three sections: \sct{sect:theory} analytically derives the method, \sct{sect:iter} and \sct{sect:pc} present two solution techniques and describes their application and development.

@@

We note that the time-dependent shape equation is similar to the time-dependent flux equation, with the following modifications:
\begin{enumerate}
\item The shape equation contains an additional term equivalent to a removal cross section,  $ \frac{1}{v^g}\frac{1}{p}\frac{dp}{dt}$.
\item The delayed neutron source term is divided by $p$.
\item The system of equations is now nonlinear due to the factorization.
\item An equation is needed to obtain the amplitude $p$.
\end{enumerate}

To derive the amplitude equation, the shape/precursors equations are weighted by a space-dependent function and integrated over the phase-space. The weight function is typically the adjoint flux $\phi^{*g}$, which can be proven to minimize truncation error \cite{duderstadt1976nuclear}.  \tcr{Check Cristian Rabitti's PhD work. There is more to it. He used a new adjoint at each time step I believe. It is interesting to read}
The final expressions are given below:
\begin{subequations}
\be
\frac{dp}{dt}=\left[\frac{\rho-\bar{\beta}}{\Lambda}\right]p+\sum_{i=1}^I\bar{\lambda}_i\xi_i
\label{eq:p}
\ee
\be
\frac{d\xi_i}{dt}=\frac{\bar{\beta}_i}{\Lambda}-\bar{\lambda}_i\xi_i \quad 1 \le i \le I 
\label{eq:c}
\ee
\end{subequations}
This equation is also known as the point reactor kinetics equation (PRKE); where the reactivity, effective delayed-neutron fraction, and delayed-neutron precursor decay constant are defined as follows:
\begin{subequations}
\be
\frac{\rho-\bar{\beta}}{\Lambda} = 
\frac{ \sum_{g=1}^G\left(\phi^{*g},\frac{\chi_p^g}{\keff}(1-\beta)\sum_{g'=1}^G \nu^{g'} \Sigma_f^{g'}\varphi^{g'} + \sum_{g'\neq g}^G\Sigma_s^{g'\to g} \varphi^{g'} -\left( -\div D^g \grad  + \Sigma_r^g \right)\varphi^g\right)}
{\sum_{g=1}^G\left(\phi^{*g},\frac{1}{v^g}\varphi^g\right)}
\label{eq:rmb}
\ee
\be
\frac{\bar{\beta}}{\Lambda} = \sum_{i=1}^I\frac{\bar{\beta}_i}{\Lambda} = 
\sum_{i=1}^I\frac{1}{\keff}\frac{\sum_{g=1}^G(\phi^{*g}, \chi_{d,i}^g\beta_i\sum_{g'=1}^G\nu^{g'} \Sigma_f^{g' }\varphi^{g'})}
{\sum_{g=1}^G\left(\phi^{*g},\frac{1}{v^g}\varphi^g\right)}
\label{eq:b}
\ee
\be
\bar{\lambda}_i = \frac{\sum_{g=1}^G(\phi^{*g},\chi_{d,i}^g\lambda_i C_i)}{\sum_{g=1}^G(\phi^{*g},\chi_{d,i}^gC_i)}
\label{eq:l}
\ee 
\end{subequations}
The following inner product definition has been used: $\left(\phi^{*g},f\right):=\int_D\phi^{*g}(\vec{r})f(\vec{r})d^3r$.\\

@@

Solving for the shape in \eqt{eq:shape} can become expensive, especially in two or three dimensions, and even more so when using the transport equations in lieu of the diffusion equations.  Using IQS, one expects the time dependence of the shape to be weaker than that of the flux itself,  thus allowing for larger time step sizes in updating the shape. The PRKE equations form a small ODE system and can be solved using a much smaller time step size. In transients where the shape varies much less than the flux, IQS can thus be very computationally effective. The two-time scale solution process, a micro scale for the PRKE and a macro scale for the shape, is illustrated in \fig{fig:iqsviz}.

@@

The major difference between the methods of these authors is the convergence criteria used.  Sissaoui and Koclas \cite{Sissaoui_1995, Koclas_1996} use fixed point iteration where the criteria is the simply the normalized difference between the last two computed shapes.  Monier in \cite{Monier_diss} also does fixed point iterations with the same criteria, except the solution is scaled by $\frac{\sum_{g=1}^G\left(\phi^{*g},\frac{1}{v^g}\varphi^g(t_{n})\right)}{\sum_{g=1}^G\left(\phi^{*g},\frac{1}{v^g}\varphi^g(t_{n+1})\right)}$ after each iteration.  Devooght in \cite{Devooght_1984} does a Newton-SOR iteration where the residual of the shape function evaluation is the convergence criteria and next iteration's solution is computed using Newton-Raphson method.\\

@@

The Predictor-Corrector (P-C) version of IQS factorizes the flux and derives the PRKE the same way as the standard version, but the solution of the coupled system of equations is different.  In the IQS P-C version, the flux equations (not the shape equations) are solved (represented by Equations~\ref{eq:flux}~and~\ref{eq:precursor}) in order to obtain a predicted flux solution. This predicted flux is then converted to a shape by rescaling it as follows:
\be
\varphi^g_{n+1} = \underbrace{\phi^g_{n+1}}_{\text{predicted}} \frac{K_0}{K_{n+1}}
\label{eq:rescale}
\ee
where the scaling factors are given by
\be
K_{n+1} =\sum_{g=1}^G\left(\phi^{*g},\frac{1}{v^g}\phi^g_{n+1}\right)
\ee
\be
K_{0} =\sum_{g=1}^G\left(\phi^{*g},\frac{1}{v^g}\varphi^g_{n+1}\right)=\sum_{g=1}^G\left(\phi^{*g},\frac{1}{v^g}\phi^g_{0} \,.\right)
\ee

@@

\tcr{is it interesting to try to converge the nonlinearities in $T$ and $\rho$ before going back to the shape?}

@@

A one-dimensional, single group FEM prototype will be created for IQS in order to test validation for simple problems. However, testing complex, real-world problems is most easily performed by implementing IQS in a FEM framework.  The Rattlesnake radiation transport framework is a excellent option for performing these problems.  Rattlesnake already contains a neutron transport method for CFEM diffusion equivalent to the brute-force method \tcr{Are you referring to SAAF? or did you mean diffusion in lieu of transport?}.  Additionally, Rattlesnake is well documented and well organized, so implementing IQS may be relatively straightforward. With this implementation, benchmark testing such as 16-A1 \cite{ANL_BPB}, TWIGL, LRA, C5G7-2D, etc. can be performed, as well as a full core simulation of the TREAT reactor.

@@

\tcr{for a paper, can we think of testing Embeded RK instead of/in addition to step-doubling?}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Possible Extensions}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Some extensions, beyond the proposed scope of this research, have been identified as follows:
